ODE and Diffusion Models of Tumor Growth

### 2024-03-14 paper.tex
Drawing
7

\documentclass{article}
\usepackage{graphicx}
\usepackage{amsmath}
\title{ODE and Diffusion Models of Tumor Growth}
\author{Daniel Henricks}

@@

Researchers have been studying cancer and the patterns behind tumor growth for thousands of years. Hippocrates
was accredited with possibly being the first to describe a tumor, and tumors were mainly described by their color
and their response to touch. However, ancient physicians did not separate benign and malignant tumors, but they
acknowledged the tumor's ability to infect nearby tissue. The application of mathematical models to tumor growth
did not start to gain traction until much later, however. In the early 20th century, one of the most famous uses of
the diffusion model was first developed: Hill released a paper discussing how the movement of lactic acid
and oyxgen throughout the body could be shaped into a diffusion problem (Hill, 1928). This paper eventually led to
Greenspan's paper which modeled the growth of tumors using diffusion, applying many of the mathematical tools employed
by Hill some years before (Greenspan, 1972). This paper will present a brief overview of the mathematics used in these
papers and will recreate some of the graphics produced as well.

@@

\section{Greenspan's paper: Models for the Growth of a Solid Tumor by Diffusion}

@@

The primary goal of Greenspan was to "infer the chemical source of growth inhibition from the most easily obtained data...the outer radius
of the nodule as a function of time." The conclusions were also described "with as little mathematics as possible" to try to attract a more
general audience, unlike the previous papers of the time.

>>>>

\documentclass{article}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{tikz}
\title{ODE and Diffusion Models of Tumor Growth}
\author{Daniel Henricks}

@@

Researchers have been studying cancer and the patterns behind tumor growth for thousands of years. Hippocrates
was accredited with possibly being the first to describe a tumor, which were mainly described by their color
and their response to touch. However, ancient physicians did not separate benign and malignant tumors, but they
acknowledged the tumor's ability to infect nearby tissue. The application of mathematical models to tumor growth
did not start to gain traction until much later, however. In the early 20th century, one of the most famous uses of
the diffusion model was first developed: Hill released a paper discussing how the movement of lactic acid
and oyxgen throughout the body could be shaped into a diffusion problem (Hill, 1928). This paper eventually led to
Greenspan's paper which modeled the growth of tumors using diffusion, applying many of the mathematical tools employed
by Hill some years before (Greenspan, 1972). This paper will present a brief overview of the mathematics used in these
papers and will recreate some of the graphics produced as well.

@@

\section{Greenspan's Paper: Models for the Growth of a Solid Tumor by Diffusion}

@@

The primary goal of Greenspan was to `infer the chemical source of growth inhibition from the most easily obtained data...the outer radius
of the nodule as a function of time.' The conclusions were also described `with as little mathematics as possible' to try to attract a more
general audience, unlike the previous papers of the time.

\subsection{Overview of the Biology}

A tumor is any abnormal group of cells, which can either be benign or malignant. Benign tumors are defined by being confined to their
original location and do not infect surrounding tissue or body sites. Malignant tumors, however, are capable of invading nearby tissue
and can spread throughout the body via metastasis. For a tumor to be classified as cancer, it must be a malignant tumor; cancers are so
dangerous because of their ability to spread to various sites in the body, making localized treatments very hard to administer.

Tumor growth is usually backed by the tumor receiving nutrients and wastes from nearby cells via a diffusion process similar to one discussed
in section 2 of the paper. When the tumor is small, most cells nearby still are receiving most of the necessary nourishment from the source of the
diffusion. In terms discussed before, the cells are absorbing $\gamma \* a$ times the normal rate of absorption, where $\gamma \in [0, 1]$ represents
the proportion of nutrients that go to the healthy cell. As the size of the tumor increases, $\gamma \rightarrow 0$; however, the growth rate of the tumor
begins to decrease as it becomes harder for it to absorb resources from the healthy cells. A necrotic core - the collection of dead and dying cells that allow
for metastasis - is formed at the center of the tumor. The steady state configuration for this tumor is typically modeled
to be a sphere with multiple layers. This leads into the variation of this model that is utilized by Greenspan.

\subsection{The Model}

The typical model used to represent a tumor is as follows:
\begin{enumerate}
    \item[(a)] The outermost layer: A layer several cells thick where cell activity proceeds as normal with exponential growth.
    \item[(b)] The middle layer: Cells are alive but do not perform mitosis.
    \item[(c)] The inner layer: This is the necrotic core. The cells here are dead or dying.
\end{enumerate}

A better illustration of this is shown in Figure \ref{fig:cross-section}:
\begin{figure}[ht]
    \centering
    \begin{tikzpicture}[scale=2]
        \draw (0, 0) circle (1.5);
        \draw (0, 0) circle (1);
        \draw (0, 0) circle (0.45);

        \draw[->, thick] (0, 0) -- (0.45, 0) node[midway, above] {$r_1$};
        \draw[->, thick] (0, 0) -- (0, 1) node[midway, right] {$r_2$};
        \draw[->, thick] (0, 0) -- (-1.07, -1.07) node[midway, above] {$r_3$};
    \end{tikzpicture}
    \caption{A cross-section of the sphere that represents the tumor. The necrotic core is situated between $0 < r < r_1$,
        the middle layer is situated between $r_1 < r < r_2$, and the healthy outer layer is located between $r_2 < r < r_3$.}
    \label{fig:cross-section}
\end{figure}

Typically, the few tumor cells can expand quite rapidly depending on several parameters, such as the cell types and nutrient levels of the body.
They can multiply in number to around a million in a growth period of days. The nodule (Figure \ref{fig:cross-section}) can expand from being microscopic
to having a radius of over 1 millimeter. Greenspan wished to apply this model to understand how $r_3$ - the outer radius of the nodule - changes
with respect to time by analyzing the steady state. To do this, several assumptions were adopted by the model:

\begin{enumerate}
    \item[(i)] The tumor is a sphere with perfect symmetry at all times, leaving the experiment to have 2 independent variables: time, $t$, and $r$, the radial distance.
    \item[(ii)] Cancer cells die when the nutrient levels, $\sigma(r, t)$ drop below a critical minimum of $\sigma_{min}.$
\end{enumerate}
